The Social Conditioning of Fear

### 2023-12-06 scripts/manuscript.qmd
edits to methods section

Our dependent variable is coded as **fear** by the GSS. Respondents were asked, "*Is there any area right around here--that is, within a mile--where you would be afraid to walk alone at night*?" During the 2022 wave, there were 3,544 answers coded as follows: 878 = yes, 1440 = no, and 1226 = NA. Notably, the GSS codes missing values with multiple indicators--".i: Inapplicable," ".n: No answer," ".d: Do not Know/Cannot Choose," and ".s: Skipped on Web." For the purposes of our preliminary study, we coded all missing data as NA. Furthermore, we applied fixed removal of NAs for our dependent variable and we ran mice imputations for all other variables to create a more robust dataset that is necessary for further modeling.

Our control and independent variables are representative of two different dimensions we sought to capture. The first dimension is individual-level or identity characteristics:

@@

-   **age**

@@

-   **sex**: Code respondent's sex

@@

Similarly, gender was a contentious variable to account for. Although the GSS has several questions aimed at capturing respondent's gender, we found them to be lacking in breathe. For example, the GSS codes **sex** within a Male/Female binary based solely on interviewer discretion. Ultimately, we opted to use **sexnow1** since it was the variable that most closely captured self-described gender. However, it remains problematic as respondents were only provided four choice: Male, Female, Transgender, and None of these. Out of 2306 cases, only 10 were coded as Transgender and 9 as None of these.

Another important variable we sought to control for was citizenship status, since citizenship status and the associated stigma of illegality and threat of deportation have been found to increase fear in daily life [@golash-boza2015; @kline2019]. However, GSS data is not conducive to controlling for this variable due to the large number of NAs. Out of 2318 cases, 2164 were coded as NAs.

The second dimension we sought to capture was community-level characteristics, i.e., the social location of individuals. For this dimension, we drew on the following variables:

@@

@@

gss_subset %>% 
  filter(!is.na(racecombo)) %>% 
  filter(racecombo != "Other") %>% 
  filter(!is.na(sexnow1)) %>% 
  filter(sexnow1 != "None of these") %>% 
  pivot_longer(
    cols = fear,
    names_to = "fear",
    values_to = "answer") %>% 
  select(racecombo, answer, sexnow1) %>% 
  mutate(fear = as.numeric(case_when(answer == "no" ~ "0",
                                 answer == "yes" ~ "1"))) %>% 
  select(racecombo, fear, sexnow1) %>% 
  group_by(racecombo, sexnow1) %>% 
  summarize(mean_fear = mean(fear)) %>% 
  pivot_wider(
    names_from = sexnow1,
    values_from = mean_fear
  ) %>% 
  kable(caption = "Social Fear by Race and Gender",
        col.names = c("Race", "Female", "Male", "Transgender"),
        digits = 3) %>% 
  kableExtra::row_spec(0:6, background = "white") 
```
@fig-FearRaceGender does a good job at plotting this relationship.

@@

```
# Discussion

\newpage
# References

>>>>

Our dependent variable is coded as **fear** by the GSS. Respondents were asked, "*Is there any area right around here--that is, within a mile--where you would be afraid to walk alone at night?*" During the 2022 wave, there were 3,544 answers coded as follows: 878 = yes, 1440 = no, and 1226 = NA. Notably, the GSS codes missing values with multiple indicators--".i: Inapplicable," ".n: No answer," ".d: Do not Know/Cannot Choose," and ".s: Skipped on Web." For the purposes of our preliminary study, we coded all missing data as NA. Furthermore, we applied fixed removal of NAs for our dependent variable and we ran mice imputations for all other variables to create a more robust dataset that is necessary for further modeling.

Our control and independent variables are representative of two different dimensions we sought to capture. The first dimension is individual-level or social identity characteristics:

@@

-   **age**: Respondent's age in numeric values.

@@

@@

Similarly, gender was a contentious variable to account for. Although the GSS has several questions aimed at capturing respondent's gender, we found them to be lacking in breadth. For example, the GSS codes **sex** within a Male/Female binary based solely on interviewer discretion. Ultimately, we opted to use **sexnow1** since it was the variable that most closely captured self-described gender. However, it remains problematic as respondents were only provided four choice: Male, Female, Transgender, and None of these. Out of 2306 cases, only 10 were coded as Transgender and 9 as None of these. Therefore, this is by no means a representative sample on gender minority population. The fact that the GSS data is primarily collected through in-person field interviews increases the possibility that minority-identifying individuals may not disclose their accurate identity information to avoid being associated with negative stereotypes.

Another important variable we sought to control for was citizenship status, since citizenship status and the associated stigma of illegality and threat of deportation have been found to increase fear in daily life [@golash-boza2015; @kline2019]. However, GSS data is not conducive to controlling for this variable due to the large number of NAs. Out of 2318 cases, 2164 were coded as NAs. In future work, we will pursue data options that allow the analysis ton include citizenship status.

The second dimension we sought to capture was community-level characteristics that would potentially affect one's sense of safety. For this dimension, we drew on the following variables:

@@

The third dimension we included were on :

@@

gss_subset %>% 
  filter(!is.na(racecombo)) %>% 
  filter(racecombo != "Other") %>% 
  filter(!is.na(sexnow1)) %>% 
  filter(sexnow1 != "None of these") %>% 
  pivot_longer(
    cols = fear,
    names_to = "fear",
    values_to = "answer") %>% 
  select(racecombo, answer, sexnow1) %>% 
  mutate(fear = as.numeric(case_when(answer == "no" ~ "0",
                                 answer == "yes" ~ "1"))) %>% 
  select(racecombo, fear, sexnow1) %>% 
  group_by(racecombo, sexnow1) %>% 
  summarize(mean_fear = mean(fear)) %>% 
  pivot_wider(
    names_from = sexnow1,
    values_from = mean_fear
  ) %>% 
  kable(caption = "Social Fear by Race and Gender",
        col.names = c("Race", "Female", "Male", "Transgender"),
        digits = 3) %>% 
  kableExtra::row_spec(0:6, background = "white") 
```

@fig-FearRaceGender does a good job at plotting this relationship.

@@

# Discussion

\newpage

# References
